---
title: "Class 5: Data Visualization"
author: "Kianna"
format: pdf
---

# Plotting in R

R has multiple plotting and graphics systems. The most popular of which is **ggplot2**.

We have already played with "base" R graphics. This comes along with R "out of the box".

```{r}
head(cars)
```

```{r}
plot(cars)
```

Compared to base R plots ggplot is much more verbose- I need to write more code to get simple plots like the one above.

To use ggplot I need to first install the ggplot2 package. To install any package in R I use the 'install.packages()' command along with the package name.

The install is a one time only requirement. The package is now on our computer. I don't need to re-install it. However, I can't just use it without loading it up with a 'library()' call.

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

All ggplot figures need at least 3 things:

-   data (this is the data.frame with our numbers)
-   aesthetics ("aes", how our data maps to the plot)
-   geoms (do want lines, points, columns, etc.)

```{r}
ggplot(data=cars) + aes(x=speed, y=dist) + geom_point() 
```

I want a trend line to show the relationship between dist and speed.

```{r}
ggplot(data=cars) + aes(x=speed, y=dist) + geom_point() + geom_line()
```

That is not what we want.

```{r}
ggplot(data=cars) + aes(x=speed, y=dist) + geom_point() + geom_smooth()
```

```{r}
bb <- ggplot(data=cars) + aes(x=speed, y=dist) + geom_point()

bb
```

```{r}
bb + geom_smooth(method = "lm", se=FALSE)
```

```{r}
bb + geom_smooth(method = "lm", se=FALSE) + 
  labs(title="Speed and Stopping Distances of Cars", 
       x="Speed (MPH)", y="Stopping Distance(ft)", 
       subtitle = "Your informative subtitle text here", 
       caption="Dataset: 'cars'")
```

  
# Gene expression example

Read the dataset from online:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

The 'head()' function will print out just the first few rows (6 by default).

```{r}
head(genes)
```

```{r}
nrow(genes)
```

```{r}
table(genes$State)
```
```{r}
round(table(genes$State)/nrow(genes) * 100,2)
```

```{r}
ggplot(genes) +aes(Condition1, Condition2, color=State) + 
  geom_point() + 
  labs(title="Some Plot", subtitle= "With a subtitle")
```

```{r}
ggplot(genes) +aes(Condition1, Condition2) + 
  geom_point() + 
  labs(title="Some Plot", subtitle= "With a subtitle")
```
```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
p
```
```{r}
p + scale_colour_manual( values=c("blue","gray","red") )
```

```{r}
p + scale_colour_manual(values=c("blue","gray","red")) +
    labs(title="Gene Expresion Changes Upon Drug Treatment",
         x="Control (no drug) ",
         y="Drug Treatment")
```


I write some text I want **bold** or *italic*
