---
title: "Halloween Candy Mini-Project"
author: "Kianna"
format: html
---

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different candy types in the dataset.

>Q2. How many fruity candy types are in the dataset?


```{r}
sum(candy[ , "fruity"])
```
There are 38 fruity candy types in this dataset.

```{r}
candy["Twix", ]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Skittles original", ]$winpercent
```

My favorite candy in the dataset is Skittles original. Its winpercent value is 63.08514%. 

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
The winpercent value for Kit Kat is 76.7686%.

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercent value for Tootsie Roll Snack Bars is 49.6535%. 

```{r}
library("skimr")
skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The histogram column looks to be on a different scale. There are no numbers in the column; it looks like an image of bars. Also, the skim_variable has no numbers in the column as it is the names of the variables themselves.

# Note: when I render the quarto document as a pdf, nothing appears in the "hist" column. However, I can see what is in the column when I open it in RStudio or render it as html. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

I think a 0 in this column means that candy type DOES NOT contain chocolate whereas a 1 means that candy type DOES contain chocolate.


>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

>Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values is not perfectly symmetrical. It represent a bell curve, but it is definitely skewed right.

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

The center of distribution lies slightly above 50% at 50.31676%.

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate <-candy$winpercent[as.logical(candy$chocolate)]
```
```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
```
```{r}
fruity <- candy$winpercent[as.logical(candy$fruity)]
```

```{r}
mean(candy$winpercent[as.logical(candy$fruity)])
```
Chocolate candy is on average ranked higher than fruity candy. The average winpercent value for chocolate candy is 60.92153% and for fruit candy is 44.11974%.

>Q12. Is this difference statistically significant?

```{r}
t.test(chocolate, fruity)
```
The reported p value when performing a two sample t-test between the winpercent values for chocolate and fruity candy was 2.871e-08. This value is below a p-value of 0.05, so we can reject the null hypothesis and conclude there is a statistically significant difference between how chocolate and fruity candy is ranked.

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent), ], n=5)
```
The five least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent), ], n=5)
```
The top 5 all time favorite candy types are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter cups.

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
```

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```


```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```


>Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

>Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.


```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```



>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )
```
The candy with the lowest price percent but highest winpercent is toosie roll midgies.


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
These are the top 5 most expensive candy types as they have the highest pricepercent values. The least populary one is Nik L Nip.

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```


```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```


>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables most anti-correlated are fruity and chocolate.

>Q23. Similarly, what two variables are most positively correlated?

The two variables most positively correlated are bar and chocolate, though it's hard to tell whether winpercent and chocolate have a higher positive correlation.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
library(plotly)
```

```{r}
ggplotly(p)
```


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The variables picked up strongly by PC1 in the positive direction are fruity, hard, and pluribus. This makes sense to me because on the correlation matrix, those variable are positive correlated with each other. Also, in my experience the fruity candy I've come across with are hard and come in a box of multiple candies, like skittles.